---
title: "11: Generalized Linear Models (ANOVA)"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2020"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
2. Apply special cases of the GLM (ANOVA) to real datasets
3. Interpret and report the results of ANOVA in publication-style formats

## Set up
```{r, message = FALSE}
getwd()
library(tidyverse)
library(agricolae)

Litter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")

# Set date to date format
Litter$collectDate <- as.Date(Litter$collectDate , format = "%Y-%m-%d")

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

### One-way ANOVA
A one-way ANOVA is the same test in practice as a two-sample t-test but for three or more groups. In R, we can  run the model with the function `lm` or `aov`, the latter of which which will allow us to run post-hoc tests to determine pairwise differences.

Does total litter biomass differ among sites at the Niwot Ridge LTER?
```{r}
# Wrangle the data
Litter.Totals <- Litter %>%
  group_by(plotID, collectDate, nlcdClass) %>%
  summarise(dryMass = sum(dryMass))

# Test for normality. 
# Note: the important assumption for GLMs is normality of residuals, 
# not necessarily normality of raw data. See below when we plot the model.
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_040"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_041"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_046"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_047"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_051"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_057"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_058"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_061"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_062"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_063"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_064"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_067"])

qqnorm(Litter.Totals$dryMass); qqline(Litter.Totals$dryMass)

# Test for equal variance
# ANOVA is robust against departures from equal variance.
bartlett.test(Litter.Totals$dryMass ~ Litter.Totals$plotID) #Low p-value, high variance, variables are not highly correlated

# Format ANOVA as aov
Litter.Totals.anova <- aov(data = Litter.Totals, dryMass ~ plotID)
summary(Litter.Totals.anova)
#High F value, and statistically significant p-value
#PlotID is a statistically significant predictor for dryMass
#Does not tell us anything about the individual plotIDs

# Format ANOVA as lm
Litter.Totals.anova2 <- lm(data = Litter.Totals, dryMass ~ plotID)
summary(Litter.Totals.anova2)
#Compares individual PlotIDs to the intercept, which in this case is plot 040
#Estiamte is the changing y intercept compared to the y intercept of plot 040
#p-value tells whether there is a significant difference

# Checking model fit and assumptions
# ANOVA is robust against departures from normality.
par(mfrow=c(2,2))
plot(Litter.Totals.anova2)
#Cooks distance helps determine outliers
#Residual vs fitted you want a horizontal line with an even cloud

# Post-hoc test
# Run this on the aov object ***
TukeyHSD(Litter.Totals.anova)
#This doesn't work too well when you have a lot of groups

# This might be a better option-
# Extract groupings for pairwise relationships
#HSD.test(aov object, predictor variable)
Litter.Totals.groups <- HSD.test(Litter.Totals.anova, "plotID", group = TRUE)
Litter.Totals.groups
#If any letter matches, it is NOT significantly different values in each group

# Graph the results
Litter.Totals.plot <- ggplot(Litter.Totals, aes(x = plotID, y = dryMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun.y = max, vjust = -1, size = 3.5,
               label = c("abc", "ab", "abcd", "bcd", "bcd", "a", 
                         "d", "abcd", "cd", "d", "abcd", "abcd")) +
  labs(x = "Plot ID", y = "Dry Mass (g)") +
  ylim(0, 85)
print(Litter.Totals.plot)
```

What information does the output give us? How might we report this information in a report?

> Overall, total litter mass differs among sites with statistical significance. (Figure 1; ANOVA, F11,198 = 4.813, p<0.0001) 
#Always report F value


### Non-parametric equivalent of ANOVA: Kruskal-Wallis Test
As with the Wilcoxon test, the Kruskal-Wallis test is the non-parametric counterpart to the one-way ANOVA. Here, the data from two or more independent samples are replaced with their ranks without regard to the grouping AND based on the between-group sum of squares calculations. 

For multiple comparisons, a p-value < 0.05 indicates that there is a significant difference between groups, but it does not indicate which groups, or in this case, months, differ from each other.

To analyze specific pairs in the data, you must use a *post hoc* test. These include the Dunn's test, a pairwise Mann-Whitney with the Bonferroni correction, or the Conover-Iman test.

```{r}
Litter.kw <- kruskal.test(Litter.Totals$dryMass ~ Litter.Totals$plotID)
Litter.kw

# There are two functions to run the Dunn Test
# dunn.test(Litter.Totals$dryMass, Litter.Totals$plotID, kw = T, 
#           table = F, list = T, method = "holm", altp = T)   #From package dunn.test
# dunnTest(Litter.Totals$dryMass, Litter.Totals$plotID)                    #From package FSA
```

## TWO-WAY ANOVA
### Main effects
A two-way ANOVA allows us to examine the effects of two categorical explanatory variables on a continuous response variable. What if we wanted to know how litter masses varied across functional groups and land cover class?

```{r}
# Format as aov
# Testing the main effects of functional group and land cover class on dry mass
# High DF means the data is very robust, a lot of data to base our analysis off of
Litter.anova.2way <- aov(data = Litter, dryMass ~ functionalGroup + nlcdClass)
summary(Litter.anova.2way)

summary(Litter$functionalGroup)
#Flowers is the intercept in this case

# Format as lm
Litter.anova.2way2 <- lm(data = Litter, dryMass ~ functionalGroup + nlcdClass)
summary(Litter.anova.2way2)

# Run a post-hoc test for pairwise differences
TukeyHSD(Litter.anova.2way)

```

### Interaction effects
We may expect the effects of functional group and land cover to be dependent on each other. For instance, we may expect needle biomass to be larger in evergreen forests than in other land types. In this case, we might choose to run an interaction effects two-way ANOVA, which will examine the individual effects of the explanatory variables as well as the interaction of the explanatory variables.

The output gives test statistics for each explanatory variable as well as the interaction effect of the explanatory variables. If the p-value for the interaction effect is less than 0.05, then we would consider the interaction among the explanatory variables to be significant.

```{r}
# Format as aov
# * represents an interaction
# If the interaction is significant, you have to leave it in the model
Litter.anova.2way3 <- aov(data = Litter, dryMass ~ functionalGroup * nlcdClass)
summary(Litter.anova.2way3)

# Format as lm
Litter.anova.2way4 <- lm(data = Litter, dryMass ~ functionalGroup * nlcdClass)
summary(Litter.anova.2way4)

```

If the interaction is significant, we interpret pairwise differences for the interaction. If the interaction is not significant, we interpret differences for the main effects only.
```{r}
# Run a post-hoc test for pairwise differences
TukeyHSD(Litter.anova.2way3)

# Call the interaction with a variable
Litter.interaction <- with(Litter, interaction(functionalGroup, nlcdClass))
Litter.anova.2way5 <- aov(data = Litter, dryMass ~ Litter.interaction)

Litter.groups <- HSD.test(Litter.anova.2way5, "Litter.interaction", group = TRUE)
Litter.groups

# Plot the results
# Edit this graph to make it more visually appealing.
# Add letters to represent statistical groupings.
Litter.anova.plot <- ggplot(Litter, aes(y = dryMass, x = functionalGroup, color = nlcdClass)) +
  geom_boxplot()+
  scale_color_manual(values=c("blue","darkblue","darkgreen")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun.y = max, vjust = -1, size = 3.5,
               label = c("d", "d", "d", "d", "d", "d", 
                        "d", "cd", "a", "b", "bc", "d", "d", "d","d", "d", "d","d", "d", "d","d", "d", "d", "d"), show.legend = FALSE) +
  labs(x = "Functional Group", y = "Dry Mass (g)")
print(Litter.anova.plot)

#Write a sentence or two about the output

```

